% Appendix A

\chapter{Sample Characteristics} % Main appendix title
\label{appendix:datasets} % For referencing this appendix elsewhere, use \ref{AppendixA}


\begin{sidewaystable}[ht]
\resizebox{\textwidth}{!}{%
\begin{tabular}{llllllllll}
\hline
\textbf{Paper} & \textbf{Diagnosis} & \textbf{CHR} & \textbf{HC} & \textbf{FEP/SZ} & \textbf{Language} & \textbf{TLC/TLI} & \textbf{Symptoms} & \textbf{Other Scales} & \textbf{Task Type} \\
\hline
\cite{nettekoven2023semantic} & CHR & 24 & 13 & 16 & English & TLI & PANSS & & \begin{tabular}[c]{@{}l@{}}picture description task, \\ story retelling\end{tabular} \\
\cite{bilgrami2022construct} & CHR & 60 & 27 & & English & TLC & SIPS/SOPS & & open‐ended interview \\
\cite{srivastava2022increased} & CHR & 172 & 126 & 65 & English & & SIPS & & open‐ended interview \\
\cite{srivastava2022p473} & CHR & 172 & 126 & 65 & English & & SIPS & & open‐ended interview \\
\cite{hitczenko2021understanding} & CHR & 36 & 41 & & English & & SIPS & WRAT & semi-structured interview \\
\cite{morgan2021natural} & CHR & 25 & 13 & 15 & English & TLI & PANSS & \begin{tabular}[c]{@{}l@{}}WRAT IQ, \\ WAIS Digit Span Test\end{tabular} & \begin{tabular}[c]{@{}l@{}}picture description task, \\ story retelling, open-ended interview\end{tabular} \\
\cite{spencer2021lower} & CHR & 24 & 13 & 16 & English & TLI & PANSS & \begin{tabular}[c]{@{}l@{}}WRAT IQ, \\ WAIS Digit Span Test\end{tabular} & \begin{tabular}[c]{@{}l@{}}picture description task, \\ story retelling, \\ open-ended interview\end{tabular} \\
\cite{haas2020linking} & CHR & 46 & 22 & & English & & SIPS/SOPS & GFS & open‐ended interview \\
\cite{rezaii2019machine} & CHR & 40 & & 23 & English & & SIPS & & semi-structured interview \\
\cite{corcoran2018prediction} & CHR & 59 & 21 & & English & K-FTDS & SIPS/SOPS & & semi-structured interview \\
\cite{gupta2018automated} & CHR & 41 & 43 & & English & & & \begin{tabular}[c]{@{}l@{}}WRAT IQ, HVLT-R, \\ LNS, CPT-IP, GFS-R\end{tabular} & written picture description \\
\cite{bedi2015automated} & CHR & 34 & 34 & & English & & SIPS/SOPS & & open‐ended interview \\
\cite{rosenstein2015language} & CHR & 30 & 25 & & English & & & & picture description task \\
\cite{argolo2023burnishing} & ARMS & 60 & 73 & & Portuguese & & SIPS & & \begin{tabular}[c]{@{}l@{}}semi-structured interview, \\ open prompt, \\ picture description task\end{tabular} \\
\cite{palominos2023coreference} & CHR & 20 & 20 & 20 & Spanish & & \begin{tabular}[c]{@{}l@{}}SIPS/SOPS \\ PANSS \end{tabular} & GAF & semi-structured interview \\
\hline
\end{tabular}
}
\caption[Sample Characteristics for Studies on Clinical High Risk Populations]{\label{tab:dataset:CHR} sample characteristics for studies on clinical high risk populations. \\ CHR - clinical high risk; ARMS - At Risk Mental States;  FEP - first episode psychosis; HC - Healthy Control; SZ - schizophrenia. TLI - Thought and Language Index; TLC - Thought, Language, and Communication Scale; K-FTDS - Kiddie Formal Thought Disorder Rating Scale. PANSS - Positive and Negative Syndrome Scale; SIPS - Structured Interview for Prodromal Symptoms; SOPS - Scale of Prodromal Symptoms. WRAT - Wide Range Achievement Test; GFS - Global Functioning Scale; LNS - Letter Number Sequencing; HVLT-R - Hopkins Verbal Learning Test – Revised; CPT-IP - Continuous Performance Test, Identical Pairs version; GFS-R - Global Functioning Scale Role; GAF - Global Assessment of Functioning.} 
\end{sidewaystable}


\begin{sidewaystable}[ht]
\resizebox{\textwidth}{!}{%
\begin{tabular}{lllllllll}
\hline
\textbf{Paper} & \textbf{Diagnosis} & \textbf{PD} & \textbf{HC} & \textbf{Language} & \textbf{TLC/TALD} & \textbf{Symptoms} & \textbf{Other Scales} & \textbf{Task Type} \\
\hline
\cite{nikzad2022does} & PD (SSD, BD, MDD) & 81 & 124 & English & TLC & SANS & BPRS & \begin{tabular}[c]{@{}l@{}}open‐ended interview, \\ picture description task\end{tabular} \\
\cite{liebenthal2022linguistic} & PD (SSD, BD, MDD) & 59 &  & English &  & PANSS &  & semi-structured interview \\
\cite{girard2022computational} & PD (SSD, BD, MDD) & 38 &  & English &  & PANSS & \begin{tabular}[c]{@{}l@{}}BPRS\\ MADRS, YMRS\end{tabular} & semi-structured interview \\
\cite{silva2022syntactic} & \begin{tabular}[c]{@{}l@{}}FEP (SZ, BD, MDD, \\ NOS, CHR)\end{tabular} & 26 & 12 & English & TLI & PANSS & SOFAS & picture description task \\
\cite{vail2018toward} & PD (SSD, BD, MDD) & 28 &  & English &  & PANSS &  & semi-structured interview \\
\cite{xu2022fully} & AVH & 134 &  & English & TALD &  & HPSVQ & story \\
\cite{xu2020centroid} & AVH & 142 &  & English & TALD &  &  & story \\
\cite{mota2022happy} & PD FEP & 24 & 33 & Portuguese &  & PANSS &  & picture description task \\
\hline
\end{tabular}
}
\caption[Sample Characteristics for Studies on Psychotic Disorder Populations]{\label{tab:dataset:PD} sample characteristics for studies on psychotic disorder populations. \\ PD - psychotic disorder; SSD - Schizophrenia Spectrum Disorder; BD - Bipolar Disorder; MDD - Major Depressive Disorder; AVH - Auditory Verbal Hallucinations; PD FEP - Psychotic Disorder First Episode Psychosis; HC - Healthy Control. TLC - Thought, Language, and Communication Scale; TALD - Thought and Language Disorder Scale. PANSS - Positive and Negative Syndrome Scale; SANS - Scale for the Assessment of Negative Symptoms; SAPS - Scale for the Assessment of Positive Symptoms. BPRS - Brief Psychiatric Rating Scale; MADRS - Montgomery-Asberg Depression Rating Scale; YMRS - Young Mania Rating Scale; SOFAS - Social and Occupational Functioning Assessment Scale; HPSVQ - Hamilton Program for Schizophrenia Voices Questionnaire.} 
\end{sidewaystable}

\begin{sidewaystable}[ht]
\resizebox{\textwidth}{!}{%
\begin{tabular}{lllllllllll}
\hline
\textbf{Paper} & \textbf{SSD} & \textbf{HC} & \textbf{Other} & \textbf{Language} & \textbf{TLC} & \textbf{Symptoms} & \textbf{Cognitive} & \textbf{Clinical} & \textbf{Other Scales} & \textbf{Task Type} \\
\hline
\cite{de2023acoustic} & 142 & 142 &  & Dutch &  & PANSS & BACS &  &  & semi-structured interview \\
\cite{voppel2023semantic} & 94 & 73 &  & Dutch &  & PANSS &  &  &  & semi-structured interview \\
\cite{corona2022assessing} & 50 & 50 &  & Dutch &  & PANSS &  &  &  & semi-structured interview \\
\cite{wouts2021belabbert} & 170 & 147 & 22 MDD & Dutch &  &  &  &  &  & semi-structured interview \\
\cite{dore2019quantification} & 50 & 50 &  & Dutch &  & PANSS &  &  &  & semi-structured interview \\
\cite{tang2023latent} & 90 & 76 &  & English & TLC & SANS &  &  &  & open‐ended interview \\
\cite{tang2022clinical} & 63 &  &  & English & TLC & SANS & WRAT-3 & BPRS & \begin{tabular}[c]{@{}l@{}}ER40, \\ AIHQ\end{tabular} & open‐ended interview \\
\cite{iter2018automatic} & 9 & 5 &  & English &  & SANS, SAPS &  &  &  & semi-structured interview \\
\cite{willits2018evidence} & 200 & 55 & HIV+ as HC & English &  &  &  &  &  & semi-structured interview \\
\cite{elvevaag2007quantifying} & 26 & 25 & \begin{tabular}[c]{@{}l@{}}10 high TD \\ 11 low TD\end{tabular} & English & TLC &  & WAIS-R IQ & BPRS &  & semi-structured interview \\
\cite{just2023validation} & 71 &  & 51 longitudinal & German &  & PANSS & Verbal IQ WST & MINI-ICF &  & semi-structured interview \\
\cite{schneider2023syntactic} & 34 & 40 & 38 MDD & German &  & SANS, SAPS & Verbal IQ WST, VLMT &  & \begin{tabular}[c]{@{}l@{}}GAF, HAM-D, \\ HAM-A\end{tabular} & picture description task \\
\cite{just2020modeling} & 20 & 20 &  & German &  & SANS, SAPS & Verbal IQ WST & CGI &  & semi-structured interview \\
\cite{ryazanskaya2020thesis} & 20 & 21 &  & Russian &  & PANSS &  &  &  & story, instruction \\
\cite{ryazanskaya2020automated} & 9 & 10 &  & Russian &  &  &  &  &  & video-description task \\
\hline
\end{tabular}
}
\caption[Sample Characteristics for Studies on Schizophrenia Spectrum Disorder Populations]{\label{tab:dataset:SDD} sample characteristics for studies on schizophrenia spectrum disorder populations. \\ SSD - Schizophrenia Spectrum Disorder; HC - Healthy Control; MDD - Major Depressive Disorder; HIV+ - Tested Positive for Human Immunodeficiency Virus; TD - Thought Disorder. TLC -  Thought, Language, and Communication Scale.  PANSS - Positive and Negative Syndrome Scale; SANS - Scale for the Assessment of Negative Symptoms; SAPS - Scale for the Assessment of Positive Symptoms. BACS - Brief Assessment of Cognition in Schizophrenia;  WRAT - Wide Range Achievement Test; WAIS - Wechsler Adult Intelligence Scale; WST - Wortschatztest; VLMT - Verbale Lern- und Merkfähigkeitstest . BPRS - Brief Psychiatric Rating Scale; MINI-ICF - Shortened version of International Classification of Functioning, Disability and Health Social Functioning Scale; CGI - Clinical Global Impression. ER40 - Penn Emotion Recognition Tests; AIHQ - Ambiguous Intentions Hostility Questionnaire; GAF - Global Assessment of Functioning; HAM-D - Hamilton Depression Scale; HAM-A - Hamilton Anxiety Scale.} 
\end{sidewaystable}


\begin{sidewaystable}[ht]
\resizebox{\textwidth}{!}{%
\begin{tabular}{lllllllllll}
\hline
\textbf{paper} & \textbf{SZ} & \textbf{HC} & \textbf{Other} & \textbf{Language} & \textbf{TLC/TLI} & \textbf{Symptoms} & \textbf{Cognitive} & \textbf{Social} & \textbf{Clinical} & \textbf{Task Type} \\
\hline
\cite{parola2022speech} & 51 & 42 &  & Chinese &  & \begin{tabular}[c]{@{}l@{}}PANSS, \\ SAPS, SANS\end{tabular} & Verbal IQ &  &  & short story \\
\cite{parola2022speech} & 111 & 129 &  & Danish &  & \begin{tabular}[c]{@{}l@{}}PANSS, \\ SAPS, SANS\end{tabular} & Verbal IQ &  &  & short story \\
\cite{voppel2021quantified} & 50 & 50 &  & Dutch &  & PANSS &  &  &  & semi-structured interview \\
\cite{jeong2023exploring} & 7 &  &  & English & TLC & SANS, SAPS &  &  &  & open‐ended interview \\
\cite{minor2023automated} & 101 &  &  & English &  & PANSS &  &  &  & semi-structured interview \\
\cite{alonso2022language} & 60 & 30 &  & English & TLI & PANSS &  & SOFAS &  & picture description task \\
\cite{alonso2022progressive} & 46 & 36 & 20, 13 longitudinal & English & TLI & PANSS & DSST, Stroop & SOFAS &  & picture description task \\
\cite{liang2022widespread} & 66 & 36 &  & English & TLI & PANSS &  & SOFAS &  & picture description task \\
\cite{aich2022towards} & 247 & 110 & 286 BD & English &  &  &  & SSPA &  & dialogue \\
\cite{tang2021natural} & 20 & 11 &  & English & TLC &  &  &  &  & open-ended interview \\
\cite{mitchell2015quantifying} &  &  &  & English &  &  &  &  &  & - \\
\cite{elvevaag2010automated} & 53 & 19 & 11 first degree relatives & English &  &  &  &  &  & heterogeneous unstructured prompts \\
\cite{parola2022speech} & 25 & 29 &  & German &  & \begin{tabular}[c]{@{}l@{}}PANSS, \\ SAPS, SANS\end{tabular} & Verbal IQ WST &  &  & short story \\
\cite{just2019coherence} & 20 & 10 &  & German &  & SANS, SAPS & Verbal IQ WST &  & CGI & semi-structured interview \\
\cite{koranova2017analyzing} & 28 &  &  & German &  & \begin{tabular}[c]{@{}l@{}}PANSS, \\ SAPS, SANS\end{tabular} &  &  &  & semi-structured interview, dialogue \\
\cite{shriki2022masking} & 23 & 28 &  & Hebrew &  & PANSS &  &  &  & written picture description, written open question \\
\cite{ziv2022morphological} & 24 & 25 &  & Hebrew &  &  &  &  &  & picture description task \\
\cite{bar2019semantic} & 24 & 27 &  & Hebrew &  & PANSS &  &  &  & written picture description, written open question \\
\cite{sarzynska2021detecting} & 35 & 35 &  & Polish & TLC &  &  &  &  & semi-structured interview \\
\cite{mota2017thought} & 21 & 21 & 20 BD & Portuguese &  & PANSS &  &  &  & story, picture description task \\
\cite{mota2016quantifying} & 21 & 21 &  & Portuguese &  & PANSS &  &  &  & short story \\
\cite{mota2014graph} & 20 & 20 & 19 BD & Portuguese &  & PANSS &  &  & BPRS & short story \\
\cite{mota2012speech} & 8 & 8 & 8 BD & Portuguese &  & PANSS &  &  & BPRS & short story \\
\cite{panicheva2019semantic}& 12 & 12 &  & Russian &  &  &  &  &  & written story \\
\hline
\end{tabular}
}
\caption[Sample Characteristics for Studies on Schizophrenia Populations]{\label{tab:dataset:SZ} sample characteristics for studies on schizophrenia populations. \\
SZ - Schizophrenia; HC - Healthy Control; BD - Bipolar Disorder. TLI - Thought and Language Index; TLC - Thought, Language, and Communication Scale.  PANSS - Positive and Negative Syndrome Scale; SANS - Scale for the Assessment of Negative Symptoms; SAPS - Scale for the Assessment of Positive Symptoms. DSST - Digit Symbol Substitution Test. SOFAS - Social and Occupational Functioning Assessment Scal; SSPA - Social Skills Performance Assessment.BPRS - Brief Psychiatric Rating Scale; CGI - Clinical Global Impression.} 
\end{sidewaystable}